---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
```{r, echo=FALSE}
htmltools::img(src= knitr::image_uri("C:/Users/prlic/Documents/GitHub/dumdum/hexagon.png"), alt = 'dumdum_hexagon', style = 'position:absolute; top:0; right:0; padding:10px; height:120px')
```


# dumdum
## Make dummy variableseasily in R
<!-- badges: start -->
![version](https://img.shields.io/badge/version-v.0.8.5-blue)
![size](https://img.shields.io/github/repo-size/prlitics/dumdum)
![issues](https://img.shields.io/github/issues/prlitics/dumdum)
<!-- badges: end -->

Dummy variables (binary, 0/1 variables) are a frequent part of analyses in the social sciences. But making them can be an arduous process in base R, which tends rely on  numerous `ifelse()`statements. This can be especially trying and headache-inducing if you have a variable with several values (e.g., race, education level, employment status, political party identification, etc.). `dumdum` cuts down on the work, gives you the ability to easily rename the new, dummied variables, and also makes [selecting reference categories](https://stats.idre.ucla.edu/other/mult-pkg/faq/general/faqwhat-is-dummy-coding/) simple. It also allows you to make dummies across many variables at once. 

There are currently a few other packages that helps users make dummy variables such as `ml`, `caret`, `tidymodels`, and `fastdummies`. What sets `dumdum` apart is that it was designed with social-scientists and non-machine learning folk in mind. Its functions are flexible and intuitive. It also requires no package dependencies; all that's necessary to have installed are the base R packages!

## Installation

You can install the latest version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("prlitics/dumdum")
```

or

```r
# install.packages("remotes")
remotes::install_github("prlitics/dumdum")
```
## Functions
There are two functions in `dumdum`
* `dummify()` which allows you to make dummy variables from a specified dataframe and variable. (This is the main function in `dumdum`)
* `dummify_across()` a wrapper for `dummify()` that lets users dummify multiple variables at once.

## Background about Example Data and dummify()

* [`dummify()` examples](#dummify)
* [`dummify_across` examples](#dummify_across)
* [Playing nice with pipes](#playing-nice-with-pipes)

These examples are going to use the [Palmer Penguins](https://github.com/allisonhorst/palmerpenguins) dataset because there are a number of "dummy-able" variables in it. (And, also, like, penguins!!)



```{r example}
library(dumdum)
penguins<-palmerpenguins::penguins
```

```{r echo=FALSE}
knitr::kable(head(penguins))
```


`dummify()` has 4 arguments; 2 mandatory and 2 options. The options are set to `NULL` as default.

```r
dummify(data, var, reference = NULL, dumNames = NULL)
```

### `data` & `var` requirements
* Currently, `dummify()` accepts data frame objects in `data`. 
* `var` accepts either the name of the column or the the column index.
* `dummify` currently accepts integer, factor, and character vectors. 

### `reference` and `dumNames` options.
* `reference` allows you to decide if you want to leave out a reference category from your values.
    * If you pass `TRUE`, it will leave out the first value it encounters.
    * If you pass the name of a value (say "Adelie"---a penguin species), it will leave that value out.
    * If you pass a vector of names (say ```r c("Adelie", "Gentoo")```), it will leave out all of the specified values.
* `dumNames` has `dummify()` rename the variables for you, so you don't have to do it afterwards.
    * The standard naming convention goes `variableName_DUM_Value`. So for a column for Adelie penguins it will be `species_DUM_Adelie`. All of the Adelie penguins will have a 1 for this variable and the other penguins will have a 0. 
    *`dumNames` can accept a character list that will rename all the columns in the order of the list (```r dumNames = c("Ad","Ge","Ch")```).
    *`dumNames` can also accept a named list so that you don't have to worry about making sure the names are in the right order. (```r dumNames = c("Ad"="Adelie","Ge"="Gentoo","Ch"="Chinstrap")```)

## Examples
### `dummify` {#dummify}
Let's say that you want to make a dummy variable for the penguins' sex because you plan to run a regression where you check to see if sex is predictive of body mass. To make dummy variables out of species, you could do this with `dummify()`:

```{r peng1, warning=FALSE}
penguins<- palmerpenguins::penguins
penguins_dummied<-dummify(data = penguins, var = "sex")
```

```{r pengview, echo=FALSE}
knitr::kable(head(penguins_dummied))
```

If you wanted to set "male" as the reference category, you could do: 
```{r warning=FALSE}
penguins_dummied<-dummify(data = penguins, var = "sex", reference = "male")
```
```{r, echo=FALSE}
knitr::kable(head(penguins_dummied))
```

The default naming convention is to make sure that the user knows what the `1` is in reference to in that column. You can also rename the columns. 

```{r warning=FALSE}
penguins_dummied<-dummify(data = penguins, var = "sex", reference = "male", dumNames = c("f","unknown"))
```
```{r, echo=FALSE}
knitr::kable(head(penguins_dummied))
```

If you didn't want to worry about putting the list of column names in the right order:
```{r warning=FALSE}
penguins_dummied<-dummify(data = penguins, var = "sex", reference = "male", dumNames = c("f"="female","unknown"=NA))
```
```{r, echo=FALSE}
knitr::kable(head(penguins_dummied))
```

### `dummify_across`
Let's say that there are multiple variables that you want to dummy across. In the case of the penguins, you might want to dummy species, as well as the island and sex. You can do so with `dummify_across()`. 

`dummify_across()` is a wrapper for `dummify()` that allows you to pass multiple variables at once. Like `dummify()`,  you specify a data frame object and you 



Currently, `dummify_across()` will only leave out the first encountered variable as a reference. 

### Playing nice with pipes
I personally am a huge fan of the [`tidyverse`](https://www.tidyverse.org/); it's what allowed me to get my feet wet with R before I could truly dive into it. I know a lot of potential `dumdum` users would also use the `tidyverse`, so it was important to me that `dumdum` functions were *pipe*-able. 

```{r}
library(magrittr)
pen_df <- penguins %>%
  dummify("sex")

```
```{r, echo=FALSE}
knitr::kable(head(pen_df))
```



## Acknowledgements
Huge thanks to [Sabrina Marasa](https://twitter.com/sabrina_marasa), who tested the package on the Mac version of R. 

## License 
This function is distributed under a [MIT license](https://choosealicense.com/licenses/mit/).

## Citation
```{r, echo=FALSE}
citation("dumdum")
```


## References
**Data & packages used in this readme.**
*  Horst AM, Hill AP, Gorman KB (2020). palmerpenguins: Palmer Archipelago (Antarctica) penguin data. R package version 0.1.0. https://allisonhorst.github.io/palmerpenguins/. doi: 10.5281/zenodo.3960218.
* Stefan Milton Bache and Hadley Wickham (2014). magrittr: A Forward-Pipe Operator for R.
  R package version 1.5. https://CRAN.R-project.org/package=magrittr

